\chapter{Research Questions and Hypothesis}
\label{ch:rq}

When speaking of a basic income the terms Unconditional Basic Income and Negative Income Tax are often used interchangeably. Following the argumentation of \cite{tondani2009} I show in section \ref{ch:experiment} that they can, in fact, be modelled in such a way that they are financially equivalent. However, as I showed in section \ref{ch:model}, prospect theory suggests that depending on their status as net-payer or net-receiver, citizens will have a preference for one or the other scheme. With the proposed experiment I intend to study this difference of choice.\\

\begin{req}\label{rq_scheme}
Will the choice between an UBI and a NIT differ significantly between participants depending on their status as net-payers or net-receivers?
\end{req}


\begin{hyp}\label{hyp:central}
Participants who are net-payers will vote for an UBI scheme, while net-receivers will prefer a NIT
\end{hyp}

\hfill \par

Further, with the introduction of a new taxation system, according to the neoclassical model of labor-leisure choice shown in section \ref{ch:optimum}, we can expect a change in the amount of work supplied. Both due to the increase in non-labor income and the taxation scheme. As shown in the same section, the effect will in theory depend on the status as net-payer or net-receiver.  

\begin{req}\label{rq_perf}
Will the change in supplied labor after the introduction of a tax-and-benefit scheme be dependant on the participant's status as net payer or net-receiver?
\end{req}


\begin{hyp}
Net payers under the NIT scheme will increase the amount of supplied labor, while net-receivers will decrease it.
\end{hyp}

But, as noted in section \ref{ch:optimum} as well, at first (before taxation) the UBI is equivalent to an increase in non-labor income without a relative change in costs between leisure and consumption. If a citizen decides on their labor supply at this point (and not after taxation) the effect should be of supply reduction across the board.

\begin{hyp}
Participants under the UBI scheme will reduce their work supplied independently of their status as net-payer or net-receiver.
\end{hyp}

